---
title: "Filtering and DimRed"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(hashmap)
library(ggplot2)
library(plotly)
library(stringr)
library(DT)
library(plyr)
library(dplyr)
library(biomaRt)
library(Biobase)
library(reshape2)
library(formattable)
library(VennDiagram)
library(hypeR)
library(xlsx)
library(ggsci)
library(ComplexHeatmap)
PATH <- ".."
```

## Recap
The summary statistics and respective plots along with eSet creation is in `00_pml_summstats_eset.{Rmd, html}`, imputation with random forest is in `00_imp_summary.{Rmd, html}`, This file contains filtering of samples along with dimensionality reduction using pca, tsne and umap.

```{r}
eSet <- readRDS(file.path(PATH, "data/2021_08_20_eset_imputed.RDS"))
table(eSet$Class)
eSet_wo_infl <- eSet[, which(pData(eSet)$Class != 'Inflammatory')]
cpm_eset <- eSet_wo_infl
exprs(cpm_eset) <- apply(exprs(cpm_eset), 2, function(x) {x/(sum(x)/1000000)})
print(dim(cpm_eset))

cpm_eset$Class <- recode(cpm_eset$Class, "Control"="1-Control", "HkNR"="2-HkNR", "Dysplasia"="3-Dysplasia", "Cancer"="4-OSCC")
cpm_eset$Class <- factor(cpm_eset$Class, levels = c("1-Control", "2-HkNR", "3-Dysplasia", "4-OSCC"))
```

## Dimensionality reduction {.tabset .tabset-fade .tabset-pills}
### PCA

```{r}
set.seed(1234)
#exprs(cpm_eset)
exprsMat <- log2(exprs(cpm_eset)+1)
rv <- matrixStats::rowVars(exprsMat)
featureSet <- order(rv, decreasing = TRUE)
exprsToPlot <- exprsMat[featureSet, , drop = FALSE]
exprsToPlot <- scale(t(exprsToPlot))
keepFeature <- (matrixStats::colVars(exprsToPlot) > 0.001)
keepFeature[is.na(keepFeature)] <- FALSE
exprsToPlot <- exprsToPlot[, keepFeature]

pca <- prcomp(exprsToPlot)
percentVar <- pca$sdev ^ 2 / sum(pca$sdev ^ 2)
pcaVariance <- data.frame(percentVar)
rownames(pcaVariance) <- colnames(pca$x)
dtp <- data.frame(pca$x)
dtp$Sample <- colnames(cpm_eset)
dtp$Class <- pData(cpm_eset)$Class
dtp$Type <-  pData(cpm_eset)$Type
pcXlab <- paste0('PC1', " ", toString(round(pcaVariance['PC1', ] * 100, 2)), "%","(Exp. Var)")
pcYlab <- paste0('PC2', " ", toString(round(pcaVariance['PC2', ] * 100, 2)), "%","(Exp. Var)")
plotly::ggplotly(ggplot(data = dtp, aes_string(x=dtp$PC1, y=dtp$PC2, label= "Sample", col = "Class", shape="Type"))+  ggplot2::geom_point() +  ggplot2::labs(x = pcXlab, y = pcYlab))
```

### tSNE

```{r}
set.seed(1234)
tsneOut <- Rtsne::Rtsne(exprsToPlot, perplexity = 5, initial_dims = max(50, ncol(cpm_eset)), max_iter = 200, pca=TRUE, pca_scale=TRUE )
tsneOut <- tsneOut$Y[, c(1, 2)]
dtp <- data.frame(tsneOut)
dtp$Sample <- colnames(cpm_eset)
dtp$Class <- pData(cpm_eset)$Class
dtp$Type <-  pData(cpm_eset)$Type


plotly::ggplotly(ggplot(data = dtp, aes_string(x=dtp$X1, y=dtp$X2, label= "Sample", col = "Class", shape="Type"))+  ggplot2::geom_point()+ ggplot2::labs(x = 'tSNE1', y = 'tSNE2'))

```

### UMAP

```{r}
ggstyle <- function(font="Helvetica", scale=1) {
  fs <- function(x) x*scale # Dynamic font scaling
  ggplot2::theme(
    plot.title = ggplot2::element_text(family=font, size=fs(26), face="bold", color="#222222"),
    plot.subtitle = ggplot2::element_text(family=font, size=fs(18), margin=ggplot2::margin(0,0,5,0)),
    plot.caption = ggplot2::element_blank(),
    legend.position = "right",
    legend.text.align = 0,
    legend.background = ggplot2::element_blank(),
    legend.title = ggplot2::element_blank(),
    legend.key = ggplot2::element_text(family=font, size=fs(20), face = "bold", color="#222222"),
    legend.text = ggplot2::element_text(family=font, size=fs(20), face = "bold", color="#222222"),
    axis.title =  ggplot2::element_text(family=font, size=fs(18), face = "bold", color="#222222"),
    axis.text = ggplot2::element_text(family=font, size=fs(18), face = "bold", color="#222222"),
    axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
    #axis.ticks = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(color="#222222"),
    panel.grid.minor = ggplot2::element_line(color="#222222"),
    panel.grid.major.y = ggplot2::element_line(color="#222222"),
    panel.grid.major.x = ggplot2::element_line(color="#222222"),
    panel.background = ggplot2::element_rect(fill="grey90"),
    strip.background = ggplot2::element_rect(fill="grey90", colour = "black"),
    strip.text = ggplot2::element_text(size=fs(22), hjust=0)
  )
}
```

## Top 500 genes UMAP
### UMAP

```{r}
set.seed(1357)
var_genes <- apply(exprs(eSet_wo_infl), 1, var)
top_2000 <- names(var_genes[order(var_genes,decreasing = TRUE)][1:500])

exprsMat <- exprs(eSet_wo_infl[top_2000])
rv <- matrixStats::rowVars(exprsMat)
featureSet <- order(rv, decreasing = TRUE)
exprsToPlot <- exprsMat[featureSet, , drop = FALSE]
exprsToPlot <- scale(t(exprsToPlot))
keepFeature <- (matrixStats::colVars(exprsToPlot) > 0.001)
keepFeature[is.na(keepFeature)] <- FALSE
exprsToPlot <- exprsToPlot[, keepFeature]

#run PCA
pca <- prcomp(exprsToPlot)
percentVar <- pca$sdev ^ 2 / sum(pca$sdev ^ 2)
pcaVariance <- data.frame(percentVar)
rownames(pcaVariance) <- colnames(pca$x)
dtp <- data.frame(pca$x)
dtp$Sample <- colnames(cpm_eset)
dtp$Class <- pData(cpm_eset)$Class
dtp$Type <-  pData(cpm_eset)$Type

custom.config <- umap::umap.defaults
custom.config$n_neighbors <- 5
custom.config$alpha <-0.1
custom.config$n_epochs <- 10000
custom.config$metric <- 'euclidean'
custom.config$init <- 'spectral'
  
umap_results <- umap::umap(dtp[,2:10], config = custom.config)
dtp <- data.frame(umap_results$layout)
dtp$Sample <- colnames(cpm_eset)
dtp$Class <- pData(cpm_eset)$Class
dtp$smoke <- as.character(pData(cpm_eset)$Smoking_status)
dtp$age <- pData(cpm_eset)$Age
dtp$sex <- pData(cpm_eset)$Sex
dtp$prog <- pData(cpm_eset)$Progression_status

p <- ggplot(data = dtp, 
             aes(x=X1,
                 y=X2, 
                 label= Sample,
                 group=Class
               )) +
  ggplot2::geom_point(size = 4, aes(shape= Class, color=Class, fill=Class)) +
  scale_shape_manual(values=c(21, 22, 23, 24))+
  scale_color_npg(alpha = 0.7)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::labs(x = 'UMAP1', y = 'UMAP2') +
  theme_bw() +
  ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                  axis.text.x = element_text(margin=ggplot2::margin(5, b=10)),
                  axis.text = element_text(size = 10, family='Helvetica', color="black"), 
                  axis.title = element_text(size = 12, family='Helvetica', color="black"),
                  legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="black")
                  )
p
  
#latest w/ progression status 
p <- dtp %>% ggplot(aes(
               x=X1, 
               y=X2, 
               label= Sample,
               group=Class, 
               shape= Class, 
               color=Class
               )) +
  ggplot2::geom_point(size = 4, aes(shape= Class, color=Class, fill=Class)) +
  scale_shape_manual(values=c(21, 22, 23, 24))+
  scale_color_npg(alpha = 0.7)+
  scale_fill_npg(alpha = 0.7)+
  geom_point(data = dtp %>% filter((prog=="Stable")), 
             pch=21,
             size=7, fill="grey",
              alpha = 0.5,
             colour="black") +
  geom_point(data = dtp %>% filter((prog=="Progressed-SCC")), 
             pch=22,
             size=7, fill="orange", 
            alpha = 0.5,
             colour="red") + 
  geom_point(data = dtp %>% filter((prog=="Progressed-Dys")), 
             pch=22,
             size=7,fill="lightblue", 
             alpha = 0.5,
              
             colour="blue") +
  ggplot2::labs(x = 'UMAP1', y = 'UMAP2') +
  theme_bw() +
  ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                  axis.title = element_blank(), 
                  legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="black"), legend.position = "bottom")
p

```

## KMeans Clustering on UMAP {.tabset .tabset-fade .tabset-pills}

```{r}
set.seed(1234)
koutput <- kmeans(umap_results$layout, 3)

cpm_eset$cluster <- koutput$cluster
#match cluster #s to already implemented in umap 
cpm_eset$cluster <- recode(cpm_eset$cluster, "3"="1", "2"="2", "1"="3")
dtp$cluster <- pData(cpm_eset)$cluster

umap_clust <- ggplot(data = dtp, aes(x=X1, y=X2, label= Sample, col = cluster, shape=Class))+
              ggplot2::geom_point(size=4, aes(shape= Class, color=cluster, fill=cluster)) + 
              scale_shape_manual(values=c(21, 22, 23, 24))+
              scale_color_manual(aesthetics = c("color", "fill"), values = c("3"="#41b6c4", "2"="#a1dab4", "1"="#fecc5c"))+
              ggplot2::labs(x = 'UMAP1', y = 'UMAP2') +
              theme_bw()+
              ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                                   axis.title = element_blank())

umap_clust

```

## Chi-sq test {.tabset .tabset-fade .tabset-pills}

```{r}
##mlcust
library(mclust)
BIC <- mclustBIC(exprsToPlot)
mod1 <- Mclust(exprsToPlot, x = BIC)
table(mod1$classification)

dtp$cluster <- mod1$classification[match(rownames(dtp),names(mod1$classification))]

ggplot(data = dtp, aes_string(x=dtp$X1, y=dtp$X2, label= "Sample", col = factor(dtp$cluster), shape='Class'))+  scale_colour_manual(values=c("red", "blue", "black")) +ggplot2::geom_point(size=3)+ ggplot2::labs(x = 'UMAP1', y = 'UMAP2')

#using mclust
clust_df <- data.frame(cluster=mod1$classification, class=cpm_eset$Class)
tab_df <- table(clust_df)
chiqs <- chisq.test(tab_df)
chiqs$residuals

#using kmeans
kmeans_df <- data.frame(cluster=koutput$cluster, class=cpm_eset$Class)
tab_df <- table(kmeans_df)
chiqs <- chisq.test(tab_df)
chiqs$residuals

```

## HVGs heatmap {.tabset .tabset-fade .tabset-pills}

```{r warning=F, message=F}
library(Biobase)
library(cba)
library(ComplexHeatmap)
library(scales)
library(circlize)
```

```{r load.data}
pml <- cpm_eset
exprs(pml) <- log2(exprs(pml)+1)
pml2K <- BS831::variationFilter(pml,ngenes=2000,score="mad",do.plot=TRUE)
```

```{r pdata.annotation}
annot <- pData(pml) %>% 
  dplyr::select(imputed_smoking_label,Sex,Progression_status,Class)  %>% 
  dplyr::rename(Smoker=imputed_smoking_label,Progression=Progression_status)  %>% 
  dplyr::mutate(Class = factor(Class, levels=c("1-Control","2-HkNR","3-Dysplasia","4-OSCC")))  %>% 
  dplyr::relocate(Class,Progression,Smoker,Sex)
```

### Heatmap {.tabset .tabset-fade .tabset-pills}

```{r ComplexHeatmap}
hm_annot <- ComplexHeatmap::HeatmapAnnotation(
  Smoker = annot$Smoker,
  Sex = annot$Sex,
  Progression = annot$Progression,
  Class = annot$Class,
  col = list(Class = c("1-Control" = "#E64B35B2",
                       "2-HkNR" = "#4DBBD5B2", 
                       "3-Dysplasia" = "#00A087B2", 
                       "4-OSCC" = "#3C5488B2"),
              Progression = c("NA"="white",
                              "Stable" = "gray",
                              "Progressed-SCC" = "red",
                              "Progressed-Dys" = "pink"), 
             Smoker = c("No"="gray", "Yes"="black"),
             Sex = c("F"="pink", "M"="lightblue")), na_col = "white")

pml2K_scaled <- BS831::scale_row(pml2K)
mad_genes <- rownames(exprs(pml2K_scaled))
```


```{r}
HALLMARK <-  msigdb_gsets("Homo sapiens", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>% strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )

krt_genes <- mad_genes[grepl(pattern = "KRT", x = mad_genes)]

epi_genes <- unique(c(HALLMARK$genesets$EPITHELIAL_MESENCHYMAL_TRANSITION[HALLMARK$genesets$EPITHELIAL_MESENCHYMAL_TRANSITION %in% rownames(exprs(pml2K_scaled))], krt_genes))
epi_index <- match(epi_genes, rownames(exprs(pml2K_scaled)))

infl_genes <- unique(c(HALLMARK$genesets$INTERFERON_GAMMA_RESPONSE[HALLMARK$genesets$INTERFERON_GAMMA_RESPONSE %in% rownames(exprs(pml2K_scaled))], HALLMARK$genesets$INFLAMMATORY_RESPONSE[HALLMARK$genesets$INFLAMMATORY_RESPONSE %in% rownames(exprs(pml2K_scaled))]))
infl_index <- match(infl_genes, rownames(exprs(pml2K_scaled)))

dend = as.dendrogram(hclust(dist(exprs(pml2K_scaled)), method = 'ward.D'))

od = order.dendrogram(dend)

selected1 = od[which(labels(dend) %in% epi_genes)]
selected2 = od[which(labels(dend) %in% infl_genes)]
```


```{r fig.width=8, fig.height=10}
ComplexHeatmap::Heatmap(Biobase::exprs(pml2K_scaled),
        name="expression", 
        top_annotation=hm_annot, 
        cluster_rows=dend,
        row_dend_reorder = F,
        cluster_columns=TRUE,
        clustering_distance_columns="euclidean",
        clustering_method_columns="ward.D", 
        column_split=3,
        show_parent_dend_line=TRUE,
        row_title="",
        show_column_names=FALSE,
        show_row_names=FALSE) +
        rowAnnotation(link1 = anno_mark(at = selected1, 
                                link_width = unit(0, "mm"),
                                link_gp = gpar(lwd=0.1),
                                padding = unit(0, "mm"),
                                labels = rownames(exprs(pml2K_scaled))[selected1], side = "right",
                                labels_gp = gpar(fontsize = 2, 
                                                 fontface = "bold",
                                                 col = "red"
                                                 )))+
        rowAnnotation(link2 = anno_mark(at = selected2, 
                                link_width = unit(0, "mm"),
                                link_gp = gpar(lwd=0.1),
                                padding = unit(0, "mm"),
                                labels = rownames(exprs(pml2K_scaled))[selected2], side = "right",
                                labels_gp = gpar(fontsize = 2, 
                                                 fontface = "bold",
                                                 col = "darkgreen"
                                                 )))  
```







