---
title: "PML diffanal"
author: "M. Muzamil Khan"
date: "2/28/2022"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/mmkhan/Documents/Research/pml/pml_wo_infl/pml_github")
```

```{r warning=FALSE, message=F}
library(ggplot2)
library(stringr)
library(DT)
library(plyr)
library(dplyr)
library(biomaRt)
library(Biobase)
library(reshape2)
library(formattable)
library(VennDiagram)
library(hypeR)
library(xlsx)
PATH <- getwd()
```

## Recap
The summary statistics and respective plots along with eSet creation is in `00_summstats_eset.{Rmd, html}`, imputation with random forest is in `01_pml_imp_summary.{Rmd, html}`,  DimRed analysis is in `02_dimred_hvg.{Rmd, html}`. This file contains differential expression analysis using DESeq2 along with significant marker info. The analysis is performed with age, sex and smoking status as covariates with histopathological group.

```{r}
eset <- readRDS(file.path(PATH, "data/2021_08_20_eset_imputed_updated.RDS"))
eSet_wo_infl <- eset
table(eSet_wo_infl$Class)
cpm_eset <- eSet_wo_infl
exprs(cpm_eset) <- apply(exprs(cpm_eset), 2, function(x) {x/(sum(x)/1000000)})
print(dim(cpm_eset))

eSet_wo_infl$Class <- recode(eSet_wo_infl$Class, "Cancer"="OSCC")
eSet_wo_infl$Class <- factor(eSet_wo_infl$Class, levels = c("Control", "HkNR", "Dysplasia", "OSCC"))
```


## Differential expression analysis (Regular) - Pairwise {.tabset .tabset-fade .tabset-pills}
### OSCC vs. Control

Reference level = Control

```{r}
cancer.vs.control <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('Control', 'OSCC'))]
colData <- data.frame(condition=cancer.vs.control$Class)
colData$sex <- cancer.vs.control$Sex
colData$smoke <- cancer.vs.control$imputed_smoking_label

#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(cancer.vs.control), 
                                      colData=colData, 
                                      design= formula(~sex+smoke+condition))

dds$condition <- relevel(dds$condition, ref = "Control")
#perform diffex
dds_results <- DESeq2::DESeq(dds)

results_cancer.vs.control<- DESeq2::results(dds_results)
summary(results_cancer.vs.control)
```

```{r}
results_cancer.vs.control <- results_cancer.vs.control[!is.na(results_cancer.vs.control$padj),]
results_cancer.vs.control$genes <- rownames(results_cancer.vs.control)
print(c("fdr<=0.05"=sum(results_cancer.vs.control$padj<=.05),
  "fdr<=0.01"=sum(results_cancer.vs.control$padj<=.01),
  "fdr<=0.001"=sum(results_cancer.vs.control$padj<=.001)))
```

```{r}
fdr <- 0.05
max_fc <- 1.5
min_fc <- 1.5
cancer.vs.control_up <- rownames(results_cancer.vs.control[which((results_cancer.vs.control$padj <= fdr) & (results_cancer.vs.control$log2FoldChange > max_fc)),])
cancer.vs.control_down <- rownames(results_cancer.vs.control[which((results_cancer.vs.control$padj <= fdr) & (results_cancer.vs.control$log2FoldChange < -min_fc)),])

c("cancer.vs.control_up"= length(cancer.vs.control_up),
  "cancer.vs.control_down"= length(cancer.vs.control_down))

cancer.vs.control_marker_list <- list(cancer.vs.control_up, cancer.vs.control_down)

DT::datatable(data.frame(results_cancer.vs.control[results_cancer.vs.control$padj<=.05 & results_cancer.vs.control$log2FoldChange >= 1.5, ]), caption = "Cancer.vs.Control-Upregulated")
DT::datatable(data.frame(results_cancer.vs.control[results_cancer.vs.control$padj<=.05 & results_cancer.vs.control$log2FoldChange <= -1.5, ]), caption = "Cancer.vs.Control-Downregulated")

```

```{r hcopt}
#function for hclust with optimal leaf ordering from S. Monti / D. Gusenleitner
hcopt <- function(d, HC=NULL, method = "ward", members = NULL){
  require("cba")
  if ( is.null(HC) ) {
    HC <- hclust(d,method=method,members=members)
  }
  #optimal leaf ordering
  ORD <- order.optimal(d,merge=HC$merge)
  HC$merge <- ORD$merge
  HC$order <- ORD$order
  HC
}
```

Significant diffex genes padj <= 0.05 and logfc +/- 1.5 with hcopt()

```{r}
topgenes <- unlist(cancer.vs.control_marker_list)
heatdata <-exprs(cancer.vs.control)[topgenes,]
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- c('Class', 'imputed_smoking_label')

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(cancer.vs.control)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```


```{r}
umap_results <- umap::umap(t(heatdata), config = umap::umap.defaults )
dtp <- data.frame(umap_results$layout)
dtp$Sample <- colnames(heatdata)
dtp$Class <- pData(cancer.vs.control)$Class
dtp$Type <-  pData(cancer.vs.control)$Type

plotly::ggplotly(ggplot(data = dtp, aes_string(x=dtp$X1, y=dtp$X2, label= "Sample", col = "Class", shape="Type")) +  
                   ggplot2::geom_point()+ ggplot2::labs(x = 'UMAP1', y = 'UMAP2'))
```


### Dysplasia vs. Control

Reference level = Control

```{r}
dys.vs.control <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('Control', 'Dysplasia'))]
colData <- data.frame(condition=dys.vs.control$Class)
colData$sex <- dys.vs.control$Sex
colData$smoke <- dys.vs.control$imputed_smoking_label

#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(dys.vs.control), 
                                      colData=colData,
                                      design = ~sex+smoke+condition)
dds$condition <- relevel(dds$condition, ref = 'Control')

#perform diffex
dds_results <- DESeq2::DESeq(dds)
#get the results in a nice tabular format
results_dys.vs.control<- DESeq2::results(dds_results)
results_dys.vs.control <- results_dys.vs.control[order(results_dys.vs.control$padj),]
#DT::datatable(as.data.frame(results_other.vs.control))
```

```{r}
results_dys.vs.control <- results_dys.vs.control[!is.na(results_dys.vs.control$padj),]
c("fdr<=0.05"=sum(results_dys.vs.control$padj<=.05),
  "fdr<=0.01"=sum(results_dys.vs.control$padj<=.01),
  "fdr<=0.001"=sum(results_dys.vs.control$padj<=.001))

```

```{r}
dys.vs.control_up <- rownames(results_dys.vs.control[which((results_dys.vs.control$padj <= fdr)&(results_dys.vs.control$log2FoldChange> +max_fc)),])
dys.vs.control_down <- rownames(results_dys.vs.control[which((results_dys.vs.control$padj <= fdr)&(results_dys.vs.control$log2FoldChange < -min_fc)),])

c("dys.vs.control_up"= length(dys.vs.control_up),
  "dys.vs.control_down"= length(dys.vs.control_down))
dys.vs.control_marker_list <- list(dys.vs.control_up, dys.vs.control_down)

DT::datatable(data.frame(results_dys.vs.control[results_dys.vs.control$padj<=.05 & results_dys.vs.control$log2FoldChange >= 1.5, ]), caption = "Dys.vs.Control-Upregulated")
DT::datatable(data.frame(results_dys.vs.control[results_dys.vs.control$padj<=.05 & results_dys.vs.control$log2FoldChange <= -1.5, ]), caption = "Dys.vs.Control-Downregulated")
```

```{r}
topgenes <- unlist(dys.vs.control_marker_list)
heatdata <- log2(exprs(dys.vs.control)[topgenes,]+1)
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- 'Class'

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(dys.vs.control)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```

### HkNR vs. Control

Reference level = Control

```{r}
hknr.vs.control <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('Control', 'HkNR'))]
colData <- data.frame(condition=hknr.vs.control$Class)
colData$sex <- hknr.vs.control$Sex
colData$smoke <- hknr.vs.control$imputed_smoking_label
#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(hknr.vs.control), 
                                      colData=colData,
                                      design = ~sex+smoke+condition)
dds$condition <- relevel(dds$condition, ref = 'Control')

#perform diffex
dds_results <- DESeq2::DESeq(dds)
#get the results in a nice tabular format
results_hknr.vs.control<- DESeq2::results(dds_results)
results_hknr.vs.control <- results_hknr.vs.control[order(results_hknr.vs.control$padj),]
#DT::datatable(as.data.frame(results_other.vs.control))
```

```{r}
results_hknr.vs.control <- results_hknr.vs.control[!is.na(results_hknr.vs.control$padj),]
c("fdr<=0.05"=sum(results_hknr.vs.control$padj<=.05),
  "fdr<=0.01"=sum(results_hknr.vs.control$padj<=.01),
  "fdr<=0.001"=sum(results_hknr.vs.control$padj<=.001))
```

```{r}
hknr.vs.control_up <- rownames(results_hknr.vs.control[which((results_hknr.vs.control$padj <= fdr)&(results_hknr.vs.control$log2FoldChange> +max_fc)),])
hknr.vs.control_down <- rownames(results_hknr.vs.control[which((results_hknr.vs.control$padj <= fdr)&(results_hknr.vs.control$log2FoldChange < -min_fc)),])

hknr.vs.control_marker_list <- list(hknr.vs.control_up, hknr.vs.control_down)

DT::datatable(data.frame(results_hknr.vs.control[results_hknr.vs.control$padj<=.05 & results_hknr.vs.control$log2FoldChange >= 1.5, ]), caption = "HkNR.vs.Control-Upregulated")
DT::datatable(data.frame(results_hknr.vs.control[results_hknr.vs.control$padj<=.05 & results_hknr.vs.control$log2FoldChange <= -1.5, ]), caption = "HkNR.vs.Control-Downregulated")

```

```{r}
topgenes <- unlist(hknr.vs.control_marker_list)
heatdata <- log2(exprs(hknr.vs.control)[topgenes,]+1)
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- 'Class'

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(hknr.vs.control)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```


### Dysplasia vs. OSCC

Reference level = Dysplasia

```{r}
dys.vs.cancer <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('OSCC', 'Dysplasia'))]
colData <- data.frame(condition=dys.vs.cancer$Class)
colData$sex <- dys.vs.cancer$Sex
colData$smoke <- dys.vs.cancer$imputed_smoking_label
#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(dys.vs.cancer), 
                                      colData=colData,
                                      design = ~sex+smoke+condition)
dds$condition <- relevel(dds$condition, ref = 'Dysplasia')

#perform diffex
dds_results <- DESeq2::DESeq(dds)
#get the results in a nice tabular format
results_dys.vs.cancer<- DESeq2::results(dds_results)
results_dys.vs.cancer <- results_dys.vs.cancer[order(results_dys.vs.cancer$padj),]
#DT::datatable(as.data.frame(results_other.vs.control))
```

```{r}
results_dys.vs.cancer <- results_dys.vs.cancer[!is.na(results_dys.vs.cancer$padj),]
c("fdr<=0.05"=sum(results_dys.vs.cancer$padj<=.05),
  "fdr<=0.01"=sum(results_dys.vs.cancer$padj<=.01),
  "fdr<=0.001"=sum(results_dys.vs.cancer$padj<=.001))
```

```{r}
dys.vs.cancer_up <- rownames(results_dys.vs.cancer[which((results_dys.vs.cancer$padj <= fdr)&(results_dys.vs.cancer$log2FoldChange> +max_fc)),])
dys.vs.cancer_down <- rownames(results_dys.vs.cancer[which((results_dys.vs.cancer$padj <= fdr)&(results_dys.vs.cancer$log2FoldChange < -min_fc)),])

dys.vs.cancer_marker_list <- list(dys.vs.cancer_up, dys.vs.cancer_down)

DT::datatable(data.frame(results_dys.vs.cancer[results_dys.vs.cancer$padj<=.05 & results_dys.vs.cancer$log2FoldChange >= 1.5, ]), caption = "Dys.vs.Cancer-Upregulated")
DT::datatable(data.frame(results_dys.vs.cancer[results_dys.vs.cancer$padj<=.05 & results_dys.vs.cancer$log2FoldChange <= -1.5, ]), caption = "Dys.vs.Cancer-Downregulated")
```

```{r}
topgenes <- unlist(dys.vs.control_marker_list)
heatdata <- log2(exprs(dys.vs.cancer)[topgenes,]+1)
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- 'Class'

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(dys.vs.cancer)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```

### HkNR vs. Cancer

Reference level = HkNR

```{r}
hknr.vs.cancer <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('OSCC', 'HkNR'))]
colData <- data.frame(condition=hknr.vs.cancer$Class)
colData$sex <- hknr.vs.cancer$Sex
colData$smoke <- hknr.vs.cancer$imputed_smoking_label
#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(hknr.vs.cancer), 
                                      colData=colData,
                                      design = ~sex+smoke+condition)
dds$condition <- relevel(dds$condition, ref = 'HkNR')

#perform diffex
dds_results <- DESeq2::DESeq(dds)
#get the results in a nice tabular format
results_hknr.vs.cancer<- DESeq2::results(dds_results)
results_hknr.vs.cancer <- results_hknr.vs.cancer[order(results_hknr.vs.cancer$padj),]
#DT::datatable(as.data.frame(results_other.vs.control))
```

```{r}
results_hknr.vs.cancer <- results_hknr.vs.cancer[!is.na(results_hknr.vs.cancer$padj),]
c("fdr<=0.05"=sum(results_hknr.vs.cancer$padj<=.05),
  "fdr<=0.01"=sum(results_hknr.vs.cancer$padj<=.01),
  "fdr<=0.001"=sum(results_hknr.vs.cancer$padj<=.001))
```

```{r}
hknr.vs.cancer_up <- rownames(results_hknr.vs.cancer[which((results_hknr.vs.cancer$padj <= fdr)&(results_hknr.vs.cancer$log2FoldChange> +max_fc)),])
hknr.vs.cancer_down <- rownames(results_hknr.vs.cancer[which((results_hknr.vs.cancer$padj <= fdr)&(results_hknr.vs.cancer$log2FoldChange < -min_fc)),])

hknr.vs.cancer_marker_list <- list(hknr.vs.cancer_up, hknr.vs.cancer_down)

DT::datatable(data.frame(results_hknr.vs.cancer[results_hknr.vs.cancer$padj<=.05 & results_hknr.vs.cancer$log2FoldChange >= 1.5, ]), caption = "HkNR.vs.Cancer-Upregulated")
DT::datatable(data.frame(results_hknr.vs.cancer[results_hknr.vs.cancer$padj<=.05 & results_hknr.vs.cancer$log2FoldChange <= -1.5, ]), caption = "HkNR.vs.Cancer-Downregulated")
```

```{r}
topgenes <- unlist(hknr.vs.control_marker_list)
heatdata <- log2(exprs(hknr.vs.cancer)[topgenes,]+1)
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- 'Class'

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(hknr.vs.cancer)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```

### HkNR vs. Dysplasia

Reference level = HkNR

```{r}
hknr.vs.dys <- eSet_wo_infl[, which(pData(eSet_wo_infl)$Class %in% c('HkNR', 'Dysplasia'))]
colData <- data.frame(condition=hknr.vs.dys$Class)
colData$sex <- as.character(hknr.vs.dys$Sex)
colData$smoke <- hknr.vs.dys$imputed_smoking_label
#make DESeq2 expected format for diffex analysis
dds <- DESeq2::DESeqDataSetFromMatrix(countData=exprs(hknr.vs.dys), 
                                      colData=colData,
                                      design = ~sex+smoke+condition)
dds$condition <- relevel(dds$condition, ref = 'HkNR')

#perform diffex
dds_results <- DESeq2::DESeq(dds)
#get the results in a nice tabular format
results_hknr.vs.dys<- DESeq2::results(dds_results)
results_hknr.vs.dys <- results_hknr.vs.dys[order(results_hknr.vs.dys$padj),]
#DT::datatable(as.data.frame(results_other.vs.control))
```

```{r}
results_hknr.vs.dys <- results_hknr.vs.dys[!is.na(results_hknr.vs.dys$padj),]
c("fdr<=0.05"=sum(results_hknr.vs.dys$padj<=.05),
  "fdr<=0.01"=sum(results_hknr.vs.dys$padj<=.01),
  "fdr<=0.001"=sum(results_hknr.vs.dys$padj<=.001))
```

```{r}
hknr.vs.dys_up <- rownames(results_hknr.vs.dys[which((results_hknr.vs.dys$padj <= fdr)&(results_hknr.vs.dys$log2FoldChange> +max_fc)),])
hknr.vs.dys_down <- rownames(results_hknr.vs.dys[which((results_hknr.vs.dys$padj <= fdr)&(results_hknr.vs.dys$log2FoldChange < -min_fc)),])

hknr.vs.dys_marker_list <- list(hknr.vs.dys_up, hknr.vs.dys_down)

DT::datatable(data.frame(results_hknr.vs.dys[results_hknr.vs.dys$padj<=.05 & results_hknr.vs.dys$log2FoldChange >= 1.5, ]), caption = "HkNR.vs.Dys-Upregulated")
DT::datatable(data.frame(results_hknr.vs.dys[results_hknr.vs.dys$padj<=.05 & results_hknr.vs.dys$log2FoldChange <= -1.5, ]), caption = "HkNR.vs.Dys-Downregulated")
```

```{r}
topgenes <- unlist(hknr.vs.dys_marker_list)
heatdata <- log2(exprs(hknr.vs.dys)[topgenes,]+1)
heatdata <- t(scale(t(heatdata)))
heatmapkey <- paste("Scaled", "counts", sep = "\n")
condition <- 'Class'

hc.row <- hcopt(as.dist(1-cor(t(heatdata))),method="ward.D")
annot_col <- pData(hknr.vs.dys)[,condition, drop=FALSE]
pheatmap::pheatmap(heatdata, cluster_rows = hc.row, show_rownames = FALSE, color = colorRampPalette(c("blue","white","red"))(100), annotation_col = annot_col)

```


## Marker info

### Pairwise Differential Markers Summary
```{r}
num_marker2 <- c("fdr<=0.05"=sum(results_cancer.vs.control$padj<=.05),
                 "fdr<=0.01"=sum(results_cancer.vs.control$padj<=.01),
                 "fdr<=0.001"=sum(results_cancer.vs.control$padj<=.001),
                 "up reg"= length(cancer.vs.control_up),
                 "down reg"= length(cancer.vs.control_down),
                 "fdr<=0.05"=sum(results_dys.vs.control$padj<=.05),
                 "fdr<=0.01"=sum(results_dys.vs.control$padj<=.01),
                 "fdr<=0.001"=sum(results_dys.vs.control$padj<=.001),
                 "up reg"= length(dys.vs.control_up),
                 "down reg"= length(dys.vs.control_down),
                 "fdr<=0.05"=sum(results_hknr.vs.control$padj<=.05),
                 "fdr<=0.01"=sum(results_hknr.vs.control$padj<=.01),
                 "fdr<=0.001"=sum(results_hknr.vs.control$padj<=.001),
                 "up reg"= length(hknr.vs.control_up),
                 "down reg"= length(hknr.vs.control_down),
                 "fdr<=0.05"=sum(results_dys.vs.cancer$padj<=.05),
                 "fdr<=0.01"=sum(results_dys.vs.cancer$padj<=.01),
                 "fdr<=0.001"=sum(results_dys.vs.cancer$padj<=.001),
                 "up reg"= length(dys.vs.cancer_up),
                 "down reg"= length(dys.vs.cancer_down),
                 "fdr<=0.05"=sum(results_hknr.vs.cancer$padj<=.05),
                 "fdr<=0.01"=sum(results_hknr.vs.cancer$padj<=.01),
                 "fdr<=0.001"=sum(results_hknr.vs.cancer$padj<=.001),
                 "up reg"= length(hknr.vs.cancer_up),
                 "down reg"= length(hknr.vs.cancer_down),
                "fdr<=0.05"=sum(results_hknr.vs.dys$padj<=.05),
                 "fdr<=0.01"=sum(results_hknr.vs.dys$padj<=.01),
                 "fdr<=0.001"=sum(results_hknr.vs.dys$padj<=.001),
                 "up reg"= length(hknr.vs.dys_up),
                 "down reg"= length(hknr.vs.dys_down)
                )
df2 <- data.frame(matrix(t(num_marker2), nrow = 6, ncol = 5, byrow = TRUE))
names(df2)<- c("fdr<=0.05", "fdr<=0.01", "fdr<=0.001", "up reg", "down reg")
rownames(df2)<- c("Cancer vs. Control", "Dysplasia vs. Control", "HkNR vs. Control",  "Dysplasia vs. Cancer", "HkNR vs. Cancer", "HkNR vs. Dysplasia")
DT::datatable(as.data.frame(df2))
pheatmap::pheatmap(df2, cluster_rows = FALSE, cluster_cols = FALSE)
```


## Diffex Results(in Excel)

```{r eval=F}
library(xlsx)

#Pairwise results
xlsx::write.xlsx(x=results_cancer.vs.control, file = file.path(PATH, "results/PML.DiffEx.results.Can.vs.Ctrl.xlsx"), row.names = TRUE)
xlsx::write.xlsx(x=results_dys.vs.control, file = file.path(PATH, "results/PML.DiffEx.results.Dys.vs.Ctrl.xlsx"), row.names = TRUE)
xlsx::write.xlsx(x=results_hknr.vs.control, file.path(PATH, "results/PML.DiffEx.results.Hknr.vs.Ctrl.xlsx"),row.names = TRUE)

xlsx::write.xlsx(x=results_dys.vs.cancer, file.path(PATH, "results/PML.DiffEx.results.Dys.vs.Can.xlsx"),row.names = TRUE)
xlsx::write.xlsx(x=results_hknr.vs.cancer, file.path(PATH, "results/PML.DiffEx.results.Hknr.vs.Can.xlsx"),row.names = TRUE)

xlsx::write.xlsx(x=results_hknr.vs.dys, file.path(PATH, "results/PML.DiffEx.results.Hknr.vs.Dys.xlsx"),row.names = TRUE)

```

Store diffanal signatures to be used in GSEA analysis.

```{r eval=F}
cancer.vs.control_sign <- list(up= cancer.vs.control_marker_list[[1]],
                              down=cancer.vs.control_marker_list[[2]])

dys.vs.control_sign <- list(up=dys.vs.control_marker_list[[1]],
                             down=dys.vs.control_marker_list[[2]])

hknr.vs.control_sign <- list(up=hknr.vs.control_marker_list[[1]],
                             down=hknr.vs.control_marker_list[[2]])

dys.vs.cancer_sign <- list(up=dys.vs.cancer_marker_list[[1]],
                             down=dys.vs.cancer_marker_list[[2]])

hknr.vs.cancer_sign <- list(up=hknr.vs.cancer_marker_list[[1]],
                             down=hknr.vs.cancer_marker_list[[2]])

hknr.vs.dys_sign <- list(up=hknr.vs.dys_marker_list[[1]],
                             down=hknr.vs.dys_marker_list[[2]])

#with control
signatures1 <- list("cancer.vs.control_up"=cancer.vs.control_marker_list[[1]],
                    "cancer.vs.control_down"= cancer.vs.control_marker_list[[2]],
                    "dys.vs.control_up"=dys.vs.control_marker_list[[1]],
                    "dys.vs.control_down"=dys.vs.control_marker_list[[2]],
                    "hknr.vs.control_up"  =hknr.vs.control_marker_list[[1]],
                    "hknr.vs.control_down"=hknr.vs.control_marker_list[[2]]
                   )

#with cancer
signatures2 <- list("dys.vs.cancer_up"=dys.vs.cancer_marker_list[[1]],
                    "dys.vs.cancer_down"=dys.vs.cancer_marker_list[[2]],
                    "hknr.vs.cancer_up"=hknr.vs.cancer_marker_list[[1]],
                    "hknr.vs.cancer_down" =hknr.vs.cancer_marker_list[[2]]
                    )


 #pairwise(infl, hknr, dys)
signatures3 <- list("hknr.vs.dys_up"=hknr.vs.dys_marker_list[[1]],
                    "hknr.vs.dys_down"=hknr.vs.dys_marker_list[[2]]
                    )

list_signs_wo_infl <- list(signatures1, signatures2, signatures3)
#saveRDS(list_signs_wo_infl, file = file.path(PATH, "results/06_22_pml_signatures_w_sex_smoke_logFC1.5_fdr0.05.RDS"))
```

