---
title: "PML hypeR and GSVA"
author: "M. Muzamil Khan"
date: "02/22/2022"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

## Recap
The summary statistics and respective plots along with eSet creation is in `00_summstats_eset.{Rmd, html}`, imputation with random forest is in `01_pml_imp_summary.{Rmd, html}`,  DimRed analysis is in `02_dimred_hvg.{Rmd, html}`. The differential expression analysis using DESeq2 along with significant marker info. is in `03_diffanalysis.{Rmd, html}`. This file contains GSVA analysis from signatures coming from relevant datasets along with hypeR GSEA from the signatures collected from differential anal results.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/mmkhan/Documents/Research/pml/pml_wo_infl/pml_github")
```

```{r warning=FALSE, message=F}
library(ggplot2)
library(stringr)
library(DT)
library(plyr)
library(dplyr)
library(biomaRt)
library(Biobase)
library(reshape2)
library(formattable)
library(VennDiagram)
library(hypeR)
library(xlsx)
library(DESeq2)
library(gridExtra)
library(ggsci)
PATH <- getwd()
```

## Load eSet {.tabset .tabset-fade .tabset-pills}

```{r}
eset <- readRDS(file.path(PATH, "data/2021_08_20_eset_imputed.RDS"))
eSet_wo_infl <- eset[, which(pData(eset)$Class != 'Inflammatory')]
table(eSet_wo_infl$Class)

eSet_wo_infl$Class <- recode(eSet_wo_infl$Class, "Cancer"="OSCC")
eSet_wo_infl$Class <- factor(eSet_wo_infl$Class, levels = c("Control", "HkNR", "Dysplasia", "OSCC"))

cpm_eset <- eSet_wo_infl
exprs(cpm_eset) <- apply(exprs(cpm_eset), 2, function(x) {x/(sum(x)/1000000)})
print(dim(cpm_eset))

cpm_eset$Class <- recode(cpm_eset$Class, "Control"="1-Control", "HkNR"="2-HkNR", "Dysplasia"="3-Dysplasia", "OSCC"="4-OSCC")
cpm_eset$Class <- factor(cpm_eset$Class, levels = c("1-Control", "2-HkNR", "3-Dysplasia", "4-OSCC"))

list_signs <- readRDS(file.path(PATH, "results/06_22_pml_signatures_w_sex_smoke_logFC1.5_fdr0.05.RDS"))
```

## Global signatures from OPMD->TCGA_HNSCC->pEMT {.tabset .tabset-fade .tabset-pills}

### OPMD signatures

```{r}
opmd <- list(up=c("SPRR2B", "DLX2", "SPRR2C", "CERS1", "CKB", "CYP19A1", "CARMIL3", "H2AC14", 
         "TUBA1B"),
         down=c("IER3", "NGEF", "TUBA4A", "ACP6", "SPIDR", "CD46", "GNPTAB", "LCA5", "ZMAT1",
         "SLC9A9", "ZNF204P", "PTCHD1", "FAM46A", "LGR5", "MUC1", "COLCA2", "DM1-AS", "ZNF418", 'NECTIN3', "MLPH", 
         "CCDC129", "TFCP2L1", "ATP6V1B1", "CRACR2B", "ERN2", "UGT1A6", "TLX1", "MUC16"))

gsva_res_opmd <-  as.data.frame(t(GSVA::gsva(exprs(cpm_eset), opmd, verbose=FALSE)))
gsva_res_opmd$diff <- gsva_res_opmd$up-gsva_res_opmd$down
```

```{r}
gsva_res_opmd$Class <- cpm_eset$Class[match(rownames(gsva_res_opmd), colnames(cpm_eset))]
opmd <- ggplot2::ggplot(gsva_res_opmd, ggplot2::aes_string(
    x = 'Class', y = 'diff', fill='Class')) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggplot2::geom_jitter(size=0.3) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    labs(y = "GSVA Score", x = "Type")+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"))

  
opmd

```

### TCGA Sigs

```{r}
tcga <- readRDS("~/Documents/Research/pml/tcga_sigs.rds")
tcga_res <- GSVA::gsva(expr = exprs(cpm_eset), gset.idx.list = tcga)
tcga_res_df <- as.data.frame(t(tcga_res))
tcga_res_df$diff <- tcga_res_df$up - tcga_res_df$down
#gsvaViolinplot(gsvaData = t(tcga_res_df), textsize = 10, eset = cpm_eset, title = 'TCGA')


tcga_res_df$Class <- cpm_eset$Class[match(rownames(tcga_res_df), colnames(cpm_eset))]
up <- ggplot2::ggplot(tcga_res_df, ggplot2::aes_string(
    x = 'Class', y = 'up', fill='Class')) +
    ggplot2::geom_boxplot() +
    ggplot2::geom_jitter(size=0.3) +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    labs(y = "GSVA Score", x = "Type")+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"))

  
up

down <- ggplot2::ggplot(tcga_res_df, ggplot2::aes_string(
    x = 'Class', y = 'down', fill='Class')) +
    ggplot2::geom_boxplot() +
    ggplot2::geom_jitter(size=0.3) +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    labs(y = "GSVA Score", x = "Type")+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"))

  
down


tcga <- ggplot2::ggplot(tcga_res_df, ggplot2::aes_string(
    x = 'Class', y = 'diff', fill='Class')) +
    ggplot2::geom_boxplot() +
    ggplot2::geom_jitter(size=0.3) +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    labs(y = "GSVA Score", x = "Type")+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"))

  
tcga


```

### pEMT Signatures

```{r}
hnsc_sign <- readRDS(file.path("~/Documents/Research/HNSC_CuratedGenesets.rds"))
hnsc_sign_epi <-  hnsc_sign[names(hnsc_sign) %in% c("Cell_Cycle_PMID29198524", "pEMT_PMID29198524", "Epithelial_Differentiation_1_PMID29198524", "Epithelial_Differentiation_2_PMID29198524", "Stress_PMID29198524", "Hypoxia_PMID29198524")]
names(hnsc_sign_epi)<- sapply(names(hnsc_sign_epi), function(x) str_split(x, pattern = "_PMID")[[1]], USE.NAMES=FALSE)[1,]
names(hnsc_sign_epi) <- recode(names(hnsc_sign_epi), "Epithelial_Differentiation_1"="Epi. Diff. 1", "Epithelial_Differentiation_2"="Epi. Diff. 2")
gsva_res_epi <- as.data.frame(t(GSVA::gsva(exprs(cpm_eset), hnsc_sign_epi, verbose=FALSE)))
gsva_res_epi$Class <- cpm_eset$Class[match(rownames(gsva_res_epi), colnames(cpm_eset))]

pemt <- ggplot2::ggplot(gsva_res_epi, ggplot2::aes_string(
    x = 'Class', y = 'pEMT', fill='Class')) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggplot2::geom_jitter(size=0.3) +
    labs(y = "GSVA Score", x = "Type")+ 
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"))


pemt
```

### Bronchial PMLs
J. Beane Bronchial PML

```{r}
pml_genes <- readRDS("~/Downloads/JBeane_PML/combined_gene_set_symbols.rds")
#names(pml_genes) <- paste("module", seq_len(length(names(pml_genes))), sep = "")
names(pml_genes) <- paste(c("module6", "module1", "module5",  "module8", "module3", "module2", "module9", "module7", "module4"), sep = "")
#selecting modules that show a positive trend
pml_genes_prolif <- pml_genes[c('module5','module9', 'module8')]

prolif_gsva <- GSVA::gsva(exprs(cpm_eset), pml_genes_prolif)
prolif_res_df <- as.data.frame(t(prolif_gsva))
prolif_res_df$Class <- cpm_eset$Class
prolif1 <- ggplot2::ggplot(prolif_res_df, ggplot2::aes_string(
    x = 'Class', y = 'module5', fill='Class')) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggplot2::geom_jitter(size=0.3) +
    labs(y = "GSVA Score", x = "Type")+ 
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"),legend.position = "none")


prolif1

prolif2 <- ggplot2::ggplot(prolif_res_df, ggplot2::aes_string(
    x = 'Class', y = 'module9', fill='Class')) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggplot2::geom_jitter(size=0.3) +
    labs(y = "GSVA Score", x = "Type")+ 
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"), legend.position = "none")


prolif2

prolif3 <- ggplot2::ggplot(prolif_res_df, ggplot2::aes_string(
    x = 'Class', y = 'module8', fill='Class')) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
    scale_fill_npg(alpha = 0.7)+
    ggplot2::geom_jitter(size=0.3) +
    labs(y = "GSVA Score", x = "Type")+ 
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    theme_bw()+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = 8),
                 axis.text.x = element_text(angle = 45, hjust = 1, margin=ggplot2::margin(5, b=10)),
                 axis.text = element_text(size = 10, family='Helvetica', color="#222222"), 
                 axis.title = element_text(size = 12, family='Helvetica', color="#222222"),
                 legend.text = ggplot2::element_text(family='Helvetica', size=(10),  color="#222222"),legend.position = "none")


prolif3
```


### CAF Signatures

```{r}
prog <- data.frame(t(exprs(cpm_eset)[c("PDGFRB", "COL1A1", "COL1A2", "COL3A1"),]))

gsvaResT <- prog
condition <- 'Class'
cond <- apply(pData(cpm_eset)[, "Class",drop = FALSE],1, paste, collapse = "_")
gsvaResT[, paste(condition, collapse = "_")] <- cond
#gsvaResT$Class <- dplyr::recode_factor(gsvaResT$Class, 'Control'='1-Control',  'HkNR'='2-HkNR', 'Dysplasia'='3-Dysplasia', 'Cancer'='4-Cancer')

gsvaResFlat <- reshape2::melt(gsvaResT, id.vars = paste(condition, collapse = "_"), variable.name = "pathway")

g1 <- ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
  x = paste(condition, collapse = "_"), y = "value",
  color = paste(condition, collapse = "_"))) +
  ggplot2::geom_boxplot() +
  scale_color_npg(alpha = 0.8)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::geom_jitter(size=0.3) +
  ggplot2::facet_wrap(~pathway, scale = "free_y",
                      ncol = nrow(prog)) +
  #ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
  ggplot2::theme(strip.text.x = ggplot2::element_text(size = 10))+
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = 'Fibroblast markers', y = "Counts(CPM)", x = "Class")+
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 10, face = "bold"))
g1
```


```{r}
pdgfb <- xlsx::read.xlsx(file = file.path(PATH, "data/pdgfb_kartha_plosone_2016.xlsx"), sheetIndex = 1)
pdgfb_genes <- pdgfb %>% dplyr::filter(Pearson.r >= 0.80)  %>% dplyr::select(Gene_ID)

pdgfb_sig <- list("pdgfb"=vapply(strsplit(pdgfb_genes$Gene_ID, "|", fixed = TRUE), "[", "", 1))
```

```{r}
pdgfb_gsva_res1 <- GSVA::gsva(expr = exprs(cpm_eset), gset.idx.list = pdgfb_sig)
```

```{r}
gsvaViolinplot <- function(gsvaData, textsize, eset, title) {
  gsvaResT <- data.frame(t(gsvaData))
  condition <- 'Class'
  cond <- apply(pData(cpm_eset)[, "Class",drop = FALSE],1, paste, collapse = "_")
  gsvaResT[, paste(condition, collapse = "_")] <- cond
  gsvaResFlat <- reshape2::melt(gsvaResT, id.vars = paste(condition,
                                                          collapse = "_"),
                                variable.name = "pathway")
  ggplot2::ggplot(gsvaResFlat, ggplot2::aes_string(
    x = paste(condition, collapse = "_"), y = "value",
    color = paste(condition, collapse = "_"))) +
    ggplot2::geom_boxplot() +
    scale_color_npg(alpha = 0.8)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::geom_jitter(size=0.3) +
    ggplot2::facet_wrap(~pathway, scale = "free_y",
                        ncol = ceiling(sqrt(nrow(gsvaData)))) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format')+
    ggplot2::theme(strip.text.x = ggplot2::element_text(size = textsize))+
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    labs(title = title, y = "GSVA Score", x = "Class")+
    theme(axis.text = element_text(size = 10), axis.title = element_text(size = 10, face = "bold"))
}

gsvaViolinplot(gsvaData = pdgfb_gsva_res1, textsize = 8,eset = cpm_eset, title = "PDGFRB sigs")
```

#### pEMT ~ pdgfr sigs

```{r}
hnsc_sign <- readRDS(file.path(PATH, "data/HNSC_CuratedGenesets.rds"))
hnsc_sign_epi <-  hnsc_sign[names(hnsc_sign) %in% c("Cell_Cycle_PMID29198524", "pEMT_PMID29198524", "Epithelial_Differentiation_1_PMID29198524", "Epithelial_Differentiation_2_PMID29198524", "Stress_PMID29198524", "Hypoxia_PMID29198524")]
names(hnsc_sign_epi)<- sapply(names(hnsc_sign_epi), function(x) str_split(x, pattern = "_PMID")[[1]], USE.NAMES=FALSE)[1,]

hnsc_pemt <- hnsc_sign_epi['pEMT']

gsva_pemt <- GSVA::gsva(expr = exprs(eSet_wo_infl), gset.idx.list = hnsc_pemt)
gsvaViolinplot(gsvaData = gsva_pemt, textsize = 8, eset = cpm_eset, title = "pEMT sigs")
```

```{r}
df_plasticity <- data.frame(t(gsva_pemt), t(pdgfb_gsva_res1), Class=eSet_wo_infl$Class)
df_plasticity$Class <- factor(df_plasticity$Class, levels = c("Control", "HkNR", "Dysplasia", "OSCC"))
```

```{r}
ggplot(df_plasticity,aes(pEMT, pdgfb)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x, ) +
   labs(y = "PDGFRb GSVA scores", x = "pEMT GSVA scores")

sp1 <- ggpubr::ggscatter(df_plasticity, x = "pEMT", y = "pdgfb",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE # Add confidence interval
   )
# Add correlation coefficient
sp1 +  ggpubr::stat_cor(method = "pearson") +
   labs(y = "PDGFRb GSVA scores", x = "pEMT GSVA scores")

sp2 <- ggpubr::ggscatter(df_plasticity, x = "pEMT", y = "pdgfb", col="Class",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE # Add confidence interval
   ) +
  scale_color_npg(alpha = 0.8)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::geom_jitter(size=0.3) 
# Add correlation coefficient
sp2 +  
  ggpubr::stat_cor(method = "pearson")  +
   labs(y = "PDGFRb GSVA scores", x = "pEMT GSVA scores")


ggplot(df_plasticity,aes(pEMT, pdgfb,  col=Class)) +
  geom_point() +
  scale_color_npg(alpha = 0.8)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::geom_jitter(size=0.3) +
  ggpubr::stat_cor(method = "pearson")+
  labs(y = "PDGFRb signature", x = "pEMT signature")

ggplot(df_plasticity,aes(pEMT, pdgfb,  col=Class)) +
  stat_summary(fun.data=mean_cl_normal) + 
  geom_smooth(method='lm', formula= y~x) +
  scale_color_npg(alpha = 0.8)+
  scale_fill_npg(alpha = 0.7)+
  ggplot2::geom_jitter(size=0.3) +
  facet_wrap(~Class, ncol = 4) +
  labs(y = "PDGFRb signature", x = "pEMT signature")

```

 
## hypeR {.tabset .tabset-fade .tabset-pills}
hypeR is run on BIOCARTA, KEGG, REACTOME separately and the .Rmd files for each is separate files. The results are stored in `0.Supporting Material>results>PML_Pathways` in the GDrive folder.

### Hallmark   

with fdr = 0.1
```{r}
rctbl_mhyp_2 <- function(mhyp,
                       show_emaps=FALSE,
                       show_hmaps=FALSE,
                       hyp_emap_args=list(top=25, val="fdr"),
                       hyp_hmap_args=list(top=25, val="fdr"), 
                       searchable = TRUE, resizable = TRUE) {

    mhyp.df <- data.frame(signature=names(mhyp$data), 
                          size=sapply(mhyp$data, function(hyp) hyp$info[["Signature Size"]]),
                          enriched=sapply(mhyp$data, function(hyp) nrow(hyp$data)),
                          gsets=sapply(mhyp$data, function(hyp) hyp$info[["Genesets"]]),
                          bg=sapply(mhyp$data, function(hyp) hyp$info[["Background"]]))
    
    tbl <- reactable(
        mhyp.df,
        showPageSizeOptions = FALSE,
        onClick = "expand",
        resizable = TRUE,
        rownames = FALSE,
        searchable = TRUE,
        filterable = TRUE,
        defaultColDef = colDef(headerClass="rctbl-outer-header"),
        columns = list(signature = colDef(name="Signature", minWidth=300),
                       size = colDef(name="Signature Size"),
                       enriched = colDef(name="Enriched Genesets"),
                       gsets = colDef(name="Genesets"),
                       bg = colDef(name="Background")),
        
        details = function(index) {
                hyp <- mhyp$data[[index]]
                rctbl_hyp(hyp, type="inner", show_emaps, show_hmaps, hyp_emap_args, hyp_hmap_args)
            },
        wrap = FALSE,
        class = "rctbl-outer-tbl",
        rowStyle = list(cursor="pointer")
        )
    
    htmltools::div(class="rctbl-outer-obj", tbl) 
}

```

```{r}
HALLMARK <-  msigdb_gsets("Homo sapiens", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>% strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )

lmultihyp1 <- hypeR(c(list_signs[[1]],list_signs[[2]], list_signs[[3]]) , HALLMARK, background = nrow(eSet_wo_infl))
rctbl_mhyp(lmultihyp1)

hyp_dots(lmultihyp1, fdr=0.1, merge = TRUE, top = 15, title = "All Pairwise")

#hyp_to_excel(lmultihyp1, file_path = "~/Documents/Research/pml/pml_wo_infl/0906_hyper.xlsx")

#for manuscript
p1 <- hyp_dots(lmultihyp1, fdr=0.1, merge = TRUE, top = 15) +
    labs(y="", x="") + theme_bw() +
    theme_bw() +
    theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")
p1


```


### Reactome   

#### Generic pathways

```{r}
REACTOME <- msigdb_gsets(species="Homo sapiens", category="C2", subcategory="CP:REACTOME", clean = TRUE)
names(REACTOME$genesets) <- names(REACTOME$genesets) %>% strsplit( "REACTOME_" ) %>% sapply( tail, 1 )

lmultihyp2 <- hypeR(c(list_signs[[1]],list_signs[[2]], list_signs[[3]]) , REACTOME, background = rownames(eSet_wo_infl))
hyp_dots(lmultihyp2, fdr=0.1, merge = TRUE,  title = "Pairwise w/ background=rownames(eset)")
rctbl_mhyp(lmultihyp2)

lmultihyp3 <- hypeR(c(list_signs[[1]],list_signs[[2]], list_signs[[3]]) , REACTOME, background = nrow(eSet_wo_infl))
hyp_dots(lmultihyp3, top = 15, fdr=0.1, merge = TRUE,  title = "Pairwise w/ background=nrow(eset)")
rctbl_mhyp(lmultihyp3)

   
# write.xlsx(lmultihyp3$data$cancer.vs.control_up$as.data.frame(), file = "~/Documents/Research/pml/pml_wo_infl/08_30_host_hyper_reactome.xlsx", sheetName = "cancer.vs.control")
# 
# for (i in 1: length(names(lmultihyp3$data))) {
#   gc()
#   write.xlsx(lmultihyp3$[names(lmultihyp3$data[i])], file = "~/Documents/Research/pml/pml_wo_infl/08_30_host_hyper_reactome.xlsx", sheetName = names(lmultihyp3$data[i]), append = T)
# }


```

#### Hierarchical

Creating heirarchical genesets 
```{r}
genesets <- REACTOME$genesets
length(genesets)
```

Clustering
```{r}
suppressPackageStartupMessages(library(hierarchicalSets))
suppressPackageStartupMessages(library(qdapTools))

mat <- genesets %>%
    qdapTools::mtabulate() %>%
    as.matrix() %>%
    t() %>%
    hierarchicalSets::format_sets()

hierarchy <- hierarchicalSets::create_hierarchy(mat, intersectLimit=1)
print(hierarchy)
```

```{r }
plot(hierarchy, type='intersectStack', showHierarchy=TRUE, label=FALSE)
plot(hierarchy, type='outlyingElements', quantiles=0.75, alpha=0.5, label=FALSE)
```



```{r }
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(stringi))
```

```{r}
find.trees <- function(d) {
    subtrees <- dendextend::partition_leaves(d)
    leaves <- subtrees[[1]]
    find.paths <- function(leaf) {
        which(sapply(subtrees, function(x) leaf %in% x))
    }
    paths <- lapply(leaves, find.paths)
    edges <- data.frame(from=c(), to=c())
    if (length(d) > 1) {
        for (path in paths) {
            for (i in seq(1, length(path)-1)) {
                edges <- rbind(edges, data.frame(from=path[i], to=path[i+1]))
            }
        }
        edges <- dplyr::distinct(edges)
        edges$from <- paste0("N", edges$from)
        edges$to <- paste0("N", edges$to)
    }
    names(subtrees) <- paste0("N", seq(1:length(subtrees)))
    nodes <- data.frame(id=names(subtrees))
    rownames(nodes) <- nodes$id
    nodes$label <- ""
    leaves <- sapply(subtrees, function(x) length(x) == 1)
    nodes$label[leaves] <- sapply(subtrees[leaves], function(x) x[[1]])
    nodes$id <- NULL
    
    # Internal nodes will not have labels, so we can generate unique hash identifiers
    ids <- stringi::stri_rand_strings(nrow(nodes), 32)
    names(ids) <- rownames(nodes)
    rownames(nodes) <- ids[rownames(nodes)]
    edges$from <- ids[edges$from]
    edges$to <- ids[edges$to]
    
    return(list("edges"=edges, "nodes"=nodes))
}
```

```{r}
trees <- lapply(hierarchy$clusters, find.trees)
length(trees)
```

Nodes
```{r}
nodes.all <- lapply(trees, function(x) x$nodes)
nodes <- do.call(rbind, nodes.all)
head(nodes)
```

Edges
```{r }
edges.all <- lapply(trees, function(x) x$edges)
edges <- data.frame(do.call(rbind, edges.all))
head(edges)
```

Create the relational genesets object
```{r}
#genesets <- hyperdb_rgsets("REACTOME", version="70.0")
rgsets_obj <- rgsets$new(genesets, nodes, edges, name="REACTOME", version=msigdb_version())
rgsets_obj

hyp1 <- hypeR(c(list_signs[[1]], list_signs[[2]], list_signs[[3]]), rgsets_obj, background = rownames(exprs(eSet_wo_infl)))
hyp_dots(hyp1, fdr=0.1, top=20, merge = TRUE, title = "Pairwise w/ background=rownames(eset)", sizes = TRUE)
rctbl_mhyp(hyp1)

#this was used in the manuscript!!!
hyp2 <- hypeR(c(list_signs[[1]], list_signs[[2]], list_signs[[3]]), rgsets_obj, background = nrow(exprs(eSet_wo_infl)))
#hyp_to_excel(hyp2, file_path = "~/Documents/Research/pml/pml_wo_infl/0906_reactome.xlsx")
rctbl_mhyp(hyp2)

hyp_dots(hyp2, fdr=0.1, top=15, merge = TRUE, title = "Pairwise w/ background=nrow(eset)", sizes = TRUE)

hyp_hmap(hyp2, fdr=0.1, top=15)

```

```{r}
#do hierarchical clust

.dots_multi_plot <- function(multihyp_data,
                             top=20,
                             abrv=50,
                             sizes=TRUE,
                             pval_cutoff=1, 
                             fdr_cutoff=1,
                             val=c("fdr", "pval"),
                             title="") {
    
    # Default arguments
    val <- match.arg(val)
    
    # Count significant genesets across signatures
    multihyp_dfs <- lapply(multihyp_data, function(hyp_obj) {
        hyp_obj$data %>%
        dplyr::filter(pval <= pval_cutoff) %>%
        dplyr::filter(fdr <= fdr_cutoff) %>%
        dplyr::select(label)
    })
    
    # Take top genesets
    labels <- names(sort(table(unlist(multihyp_dfs)), decreasing=TRUE))
    if (!is.null(top)) labels <- head(labels, top)
    
    # Handle empty dataframes
    if (length(labels) == 0) return(ggempty())
    
    # Create a multihyp dataframe
    dfs <- lapply(multihyp_data, function(hyp_obj) {
        hyp_df <- hyp_obj$data
        hyp_df[hyp_df$label %in% labels, c("label", val), drop=FALSE]
    })
    
    df <- suppressWarnings(Reduce(function(x, y) merge(x, y, by="label", all=TRUE), dfs))
    colnames(df) <- c("label", names(dfs))
    rownames(df) <- df$label
    df <- df[rev(labels), names(dfs)]
    
    # Abbreviate labels
    label.abrv <- substr(rownames(df), 1, abrv)
    if (any(duplicated(label.abrv))) {
        stop("Non-unique labels after abbreviating")
    } else {
        rownames(df) <- factor(label.abrv, levels=label.abrv)   
    }
    
    if (val == "pval") {
        cutoff <- pval_cutoff
        color.label <- "P-Value"
    }
    if (val == "fdr") {
        cutoff <- fdr_cutoff
        color.label <- "FDR"
    }
    
    df.melted <- reshape2::melt(as.matrix(df))
    colnames(df.melted) <- c("label", "signature", "significance")
    df.melted$size <- if(sizes) df.melted$significance else 1
    return(df.melted)
}

```

arrange the dotplot according to clustered groups

```{r}
.reverselog_trans <- function(base=exp(1)) {
    trans <- function(x) -log(x, base)
    inv <- function(x) base^(-x)
    scales::trans_new(paste0("reverselog-", format(base)), trans, inv, 
              scales::log_breaks(base=base), 
              domain=c(1e-100, Inf))
}

df <- .dots_multi_plot(hyp2$data, top = 15, abrv = 75, sizes = TRUE, fdr_cutoff = 0.1, pval_cutoff = 0.05, val = "fdr", title)
df <- df[df$significance <= 0.1, ]
#rownames(df) <- NULL
df %>%
    ggplot(aes(x=signature, y=(factor(label)), color=significance, size=size)) +
    geom_point() +
    scale_color_continuous(low="#114357", high="#E53935", trans=.reverselog_trans(10)) +
    scale_size_continuous(trans=.reverselog_trans(10), guide="none") +
    theme_bw() + 
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))


label_ordered <- c("Extracellular Matrix Organization", "Activation Of Matrix Metalloproteinases", "Degradation Of The Extracellular Matrix", "Cytokine Signaling In Immune System", "Interleukin 4 And Interleukin 13 Signaling", "Anti Inflammatory Response Favouring Leishmania Parasite Infection", "Fcgr Activation",  "Fcgr3a Mediated Il10 Synthesis",  "Immunoregulatory Interactions Between A Lymphoid And A Non Lymphoid Cell", "Formation Of The Cornified Envelope", "Collagen Formation", "Collagen Degradation", "Biological Oxidations", "Fatty Acids", "Cytochrome P450 Arranged By Substrate Type")
df <- df[order(match(df$label, label_ordered)),]

df$label <- factor(df$label, levels = rev(label_ordered))

df %>%
    ggplot(aes(x=signature, y=(factor(label)), color=significance, size=size)) +
    geom_point() +
    scale_color_continuous(low="#114357", high="#E53935", trans=.reverselog_trans(10)) +
    scale_size_continuous(trans=.reverselog_trans(10), guide="none") +  theme_bw() 
# +theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")

#for manuscript
p1 <- df %>%
    ggplot(aes(x=signature, y=(factor(label)), color=significance, size=size)) +
    geom_point() +
    scale_color_continuous(low="#114357", high="#E53935", trans=.reverselog_trans(10)) +
    scale_size_continuous(trans=.reverselog_trans(10), guide="none") +  theme_bw() +
    theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")

p1


```


### KEGG   

with fdr = 0.1

```{r}
KEGG <- msigdb_gsets(species="Homo sapiens", category = 'C2', subcategory = 'CP:KEGG')
#enrichr_gsets("KEGG_2019_Human")
lmultihyp1 <- hypeR(list_signs[[1]], KEGG)
hyp_dots(lmultihyp1, fdr=0.1, merge = TRUE, title = "PML with Controls")
rctbl_mhyp(lmultihyp1)

lmultihyp2 <- hypeR(list_signs[[2]], KEGG)
hyp_dots(lmultihyp2, fdr=0.1, merge = TRUE, title = "PML with Cancer")
rctbl_mhyp(lmultihyp2)

lmultihyp3 <- hypeR(list_signs[[3]], KEGG)
hyp_dots(lmultihyp3, fdr=0.1, merge = TRUE, title = "Pairwise PML")
rctbl_mhyp(lmultihyp3)

```

### GO 

with fdr = 0.1

```{r}
GO <- msigdb_gsets("Homo sapiens", "C5", subcategory = 'CC')
names(GO$genesets) <- names(GO$genesets) %>% strsplit( "GOMF_" ) %>% sapply( tail, 1 )
length(unique(names(GO$genesets)))
lmultihyp1 <- hypeR(list_signs[[1]], GO)
hyp_dots(lmultihyp1, fdr=0.1, merge = TRUE, title = "PML with Controls")
rctbl_mhyp(lmultihyp1)

lmultihyp2 <- hypeR(list_signs[[2]], GO)
hyp_dots(lmultihyp2, fdr=0.1, merge = TRUE, title = "PML with Cancer")
rctbl_mhyp(lmultihyp2)

lmultihyp3 <- hypeR(list_signs[[3]], GO)
hyp_dots(lmultihyp3, fdr=0.1, merge = TRUE, title = "Pairwise PML")
rctbl_mhyp(lmultihyp3)


lmultihyp1 <- hypeR(c(list_signs[[1]],list_signs[[2]], list_signs[[3]]) , GO, fdr = 0.1)
hyp_dots(lmultihyp1, fdr=0.1, merge = TRUE, top = 50, title = "All Pairwise", abrv = 50) 
rctbl_mhyp(lmultihyp1)
```




