---
title: "PML Microbiome Analysis w/o Infl"
author: "M. Muzamil Khan"
date: "06/23/2022"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  html_notebook:
    code_folding: hide
    theme: flatly
    toc: yes
    toc_float: true
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/mmkhan/Documents/Research/pml/pml_wo_infl/pml_github")
```

```{r warning=FALSE, message=F}
library(ggplot2)
library(stringr)
library(DT)
library(plyr)
library(dplyr)
library(biomaRt)
library(Biobase)
library(reshape2)
library(formattable)
library(VennDiagram)
library(hypeR)
library(xlsx)
library(animalcules)
library(magrittr)
library(SummarizedExperiment)
library(ggsci)
library(eply)
library(igraph)
PATH <- getwd()
```

# Microbiome Analysis 

Microbiome analysis of oral PMLs and cancer from total trx sequencing. Pathoscope was used to filter, align and map the reads to several organisms including bacterial, viral and fungi.

## load host eset {.tabset .tabset-fade .tabset-pills}

Load eset

```{r}
eset <- readRDS(file.path(PATH, "data/2021_08_20_eset_imputed.RDS"))
eSet_wo_infl <- eset[, which(pData(eset)$Class != 'Inflammatory')]
table(eSet_wo_infl$Class)

eSet_wo_infl$Class <- recode(eSet_wo_infl$Class, "Cancer"="OSCC")
eSet_wo_infl$Class <- factor(eSet_wo_infl$Class, levels = c("Control", "HkNR", "Dysplasia", "OSCC"))

cpm_eset <- eSet_wo_infl
exprs(cpm_eset) <- apply(exprs(cpm_eset), 2, function(x) {x/(sum(x)/1000000)})
print(dim(cpm_eset))

cpm_eset$Class <- recode(cpm_eset$Class, "Control"="1-Control", "HkNR"="2-HkNR", "Dysplasia"="3-Dysplasia", "OSCC"="4-OSCC")
cpm_eset$Class <- factor(cpm_eset$Class, levels = c("1-Control", "2-HkNR", "3-Dysplasia", "4-OSCC"))
```


## load animalcules MAE {.tabset .tabset-fade .tabset-pills}
New animalcules file w/o infl

File created from the shiny app with run_animalcules() easy-to-upload .tsv files from pathoscope. 

```{r warning=F}
MAE <- readRDS(file.path(PATH, "data/animalcules_data_2022-03-14.rds"))

colData(MAE[['MicrobeGenetics']]) <- cbind(colData(MAE[['MicrobeGenetics']]), "Sample_ID"=rownames(colData(MAE[['MicrobeGenetics']])))
    
#add smoking and progression status to colData
colData(MAE[['MicrobeGenetics']]) <- cbind(colData(MAE[['MicrobeGenetics']]), pData(cpm_eset)[match(MAE[['MicrobeGenetics']]$Sample_ID, gsub(cpm_eset$Sample_ID, pattern = "_", replacement = "-")), c('Smoking_status', 'imputed_smoking_label', 'Progression_status')])
    
MAE[['MicrobeGenetics']]$Progression_status <- ifelse(is.na(MAE[['MicrobeGenetics']]$Progression_status) | (MAE[['MicrobeGenetics']]$Progression_status =="Stable"), MAE[['MicrobeGenetics']]$Class, MAE[['MicrobeGenetics']]$Progression_status)

p1 <- animalcules::relabu_barplot(MAE,
                    tax_level="phylum",
                    sort_by= "conditions",
                    sample_conditions=c('Class'),
                    show_legend=TRUE)
p1                 

p2 <- animalcules::relabu_barplot(MAE,
                    tax_level="genus",
                    sort_by= "conditions",
                    sample_conditions=c('Class'),
                    show_legend=TRUE)
p2

p3 <- animalcules::relabu_barplot(MAE,
                    tax_level="species",
                    sort_by= "conditions",
                    sample_conditions=c('Class'),
                    show_legend=TRUE)
p3

```


## Wrappers

```{r}
## creating a relabu barplots wrapper to rotate the axix but didn't work
relabu_wrapper <- function (MAE, tax_level, order_organisms = c(), sort_by = c("nosort", 
  "conditions", "organisms", "alphabetically"), group_samples = FALSE, 
  group_conditions = "ALL", sample_conditions = c(), isolate_samples = c(), 
  discard_samples = c(), show_legend = TRUE) 
{
  sort_by <- match.arg(sort_by)
  MAE <- mae_pick_samples(MAE, isolate_samples, discard_samples)
  microbe <- MAE[["MicrobeGenetics"]]
  tax_table <- as.data.frame(rowData(microbe))
  sam_table <- as.data.frame(colData(microbe))
  counts_table <- as.data.frame(assays(microbe))[, rownames(sam_table)]
  sam_table %<>% df_char_to_factor()
  relabu_table <- counts_table %>% upsample_counts(tax_table, 
    tax_level) %>% counts_to_relabu() %>% base::t() %>% 
    base::as.data.frame()
  if (group_samples & !is.null(group_conditions)) {
    if (group_conditions == "ALL") {
      relabu_table$covariate <- rep("ALL", nrow(relabu_table))
    }
    else {
      relabu_table$covariate <- sam_table[[group_conditions]]
    }
    relabu_table <- relabu_table %>% reshape2::melt(id.vars = "covariate") %>% 
      S4Vectors::aggregate(. ~ variable + covariate, ., 
        mean) %>% reshape2::dcast(formula = covariate ~ 
      variable) %>% magrittr::set_rownames(.[["covariate"]]) %>% 
      dplyr::select(-one_of(c("covariate")))
    sam_table <- rownames(relabu_table) %>% as.data.frame() %>% 
      magrittr::set_colnames(c(group_conditions)) %>% 
      magrittr::set_rownames(rownames(relabu_table))
  }
  relabu_table <- relabu_table[, order(colSums(relabu_table)), 
    drop = FALSE]
  if (!is.null(order_organisms)) {
    org_order <- c(setdiff(colnames(relabu_table), order_organisms), 
      rev(order_organisms))
    relabu_table <- relabu_table[, org_order]
  }
  if (sort_by == "alphabetically") {
    org_order <- sort(colnames(relabu_table), decreasing = TRUE)
    relabu_table <- relabu_table[, org_order]
  }
  if (sort_by == "organisms") {
    for (i in seq_len(ncol(relabu_table))) {
      relabu_table <- relabu_table[order(relabu_table[, 
        i]), ]
    }
  }
  if (!is.null(sample_conditions) || (group_samples && group_conditions != 
    "ALL")) {
    if (!group_samples) {
      sam_table <- sam_table[, sample_conditions, drop = FALSE]
    }
    if (sort_by == "conditions") {
      for (i in ncol(sam_table):1) {
        sam_table <- sam_table[order(sam_table[[i]]), 
          , drop = FALSE]
      }
      relabu_table <- relabu_table[order(match(rownames(relabu_table), 
        rownames(sam_table))), , drop = FALSE]
    }
    else {
      sam_table <- sam_table[order(match(rownames(sam_table), 
        rownames(relabu_table))), , drop = FALSE]
    }
    if (nrow(sam_table) > 1) {
      hover.txt <- c()
      for (i in seq_len(ncol(sam_table))) {
        hover.txt <- cbind(hover.txt, as.character(sam_table[[i]]))
      }
      mat <- sam_table %>% data.matrix() %>% apply(2, 
        function(x) (x - min(x))/(max(x) - min(x)))
      hm <- plotly::plot_ly(x = colnames(mat), y = rownames(mat), 
        z = mat, type = "heatmap", showscale = FALSE,
        hoverinfo = "x+y+text", text = hover.txt) %>% 
        layout(xaxis = list(title = "", tickangle = -45), 
          yaxis = list(showticklabels = FALSE, type = "category", 
            ticks = ""))
    }
  }
  relabu_table$samples <- rownames(relabu_table)
  sbp <- plotly::plot_ly(relabu_table, y = ~samples, x = relabu_table[[colnames(relabu_table)[1]]], 
    type = "bar", textposition = "outside", orientaton="h",
    name = substr(colnames(relabu_table)[1], 1, 40)) %>% 
    layout(font = list(size = 10), xaxis = list(title = "Relative Abundance", 
      automargin = TRUE), yaxis = list(title = "", type = "category", 
      tickmode = "array", tickvals = rownames(relabu_table), 
      showticklabels = FALSE, categoryorder = "trace", 
      automargin = TRUE), barmode = "stack", showlegend = show_legend)
  for (i in 2:(ncol(relabu_table) - 1)) {
    sbp <- add_trace(sbp, x = relabu_table[[colnames(relabu_table)[i]]], 
      name = substr(colnames(relabu_table)[i], 1, 40))
  }
  if (exists("hm") && nrow(sam_table) > 1) {
    hm_sbp <- subplot(hm, sbp, widths = c(0.1, 0.9))
    hm_sbp$elementId <- NULL
    return(hm_sbp)
  }
  else {
    sbp$elementId <- NULL
    return(sbp)
  }
}


relabu_boxplot_wrapper <- function (MAE, tax_level, condition, organisms = c(), datatype = c("counts", 
  "relative abundance", "logcpm")) 
{
  datatype <- match.arg(datatype)
  microbe <- MAE[["MicrobeGenetics"]]
  tax_table <- as.data.frame(rowData(microbe))
  sam_table <- as.data.frame(colData(microbe))
  counts_table <- as.data.frame(assays(microbe))[, rownames(sam_table)]
  sam_table %<>% df_char_to_factor()
  df <- counts_table %>% upsample_counts(tax_table, tax_level) %>% 
    {
      if (datatype == "relative abundance") {
        animalcules::counts_to_relabu(.)
      }
      else if (datatype == "logcpm") {
        animalcules::counts_to_logcpm(.)
      }
      else {
        .
      }
    } %>% .[organisms, , drop = FALSE] %>% t() %>% as.data.frame() %>% 
    merge(sam_table[, condition, drop = FALSE], by = 0, 
      all = TRUE) %>% reshape2::melt(by = organisms, variable.name = "organisms")
  
   g <- ggplot2::ggplot(df, 
                       ggplot2::aes(x = organisms, y = value, color = Class, fill = Class)) + 
    geom_boxplot() + 
    viridis::scale_color_viridis(alpha = 0.90, discrete=T)+
    viridis::scale_fill_viridis(alpha = 0.90, discrete =T)+
    ggplot2::geom_jitter(size=0.2) +
    theme_bw() + labs(y = "log-CPM")+
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  return(g)
}

```

## Diversity plots

```{r}
alpha_diversity_wrapper <- function (MAE, tax_level, condition, alpha_metric = c("inverse_simpson", 
  "gini_simpson", "shannon", "fisher", "coverage", "unit")) 
{
  microbe <- MAE[["MicrobeGenetics"]]
  tax_table <- as.data.frame(SummarizedExperiment::rowData(microbe))
  sam_table <- as.data.frame(SummarizedExperiment::colData(microbe))
  counts_table <- as.data.frame(SummarizedExperiment::assays(microbe))[, rownames(sam_table)]
  counts_table %<>% upsample_counts(tax_table, tax_level)
  sam_table$richness <- diversities(counts_table, index = alpha_metric)
  colnames(sam_table)[ncol(sam_table)] <- "richness"
  colnames(sam_table)[which(colnames(sam_table) == condition)] <- "condition"
  
  g <- ggplot2::ggplot(sam_table, 
                       ggplot2::aes(condition, richness, color = condition, fill = condition)) + 
    geom_boxplot() + 
    viridis::scale_color_viridis(alpha = 0.90, discrete=TRUE)+
    viridis::scale_fill_viridis(alpha = 0.90, discrete = TRUE)+
    ggplot2::geom_jitter(size=0.3) +
    ggpubr::stat_compare_means(method = "anova",  label = 'p.format') +
    theme_bw() +
    ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggplot2::labs(title = paste("Alpha diversity between ", 
                                condition, " (", alpha_metric, ")", sep = ""))
    g <- g + labs(y = alpha_metric)
    return(g)
}

```

```{r}
animalcules::alpha_div_boxplot(MAE = MAE, tax_level = "genus", condition = "Class", alpha_metric = "shannon")
animalcules::diversity_beta_heatmap(MAE = MAE, tax_level = "genus", input_beta_method = "bray", input_bdhm_select_conditions =   c("Class"), input_bdhm_sort_by = "conditions")
```

```{r warning=F}
suppressPackageStartupMessages(library(plotly))

animalcules::alpha_div_boxplot(MAE = MAE, tax_level = "genus", condition = "Class", alpha_metric = "shannon")
animalcules::alpha_div_boxplot(MAE = MAE, tax_level = "species", condition = "Class", alpha_metric = "shannon")

animalcules::diversity_beta_heatmap(MAE = MAE, tax_level = "genus", input_beta_method = "bray",input_bdhm_select_conditions =   "Class", input_bdhm_sort_by = "conditions")
animalcules::diversity_beta_heatmap(MAE = MAE, tax_level = "species", input_beta_method = "bray",input_bdhm_select_conditions =   "Class", input_bdhm_sort_by = "conditions")
```

## Rel abu of MoI {.tabset .tabset-fade .tabset-pills}

```{r}
relabu_boxplot_wrapper(MAE = MAE, tax_level = "genus", condition = "Class", organisms = c("Fusobacterium", "Neisseria", "Prevotella", "Shewanella", "Streptococcus", "Candida"), datatype = "logcpm")
```

## Dim. Reduction {.tabset .tabset-fade .tabset-pills}

```{r}
animalcules::dimred_umap(MAE = MAE,tax_level = "genus", color = "Class", n_neighbors = 5, n_epochs = 1000, cx = 1, cy = 2, cz = 3)

animalcules::dimred_pcoa(MAE = MAE,tax_level = "genus", color = "Class", method = "jaccard", axx = 1, axy = 2, axz = 3)
```

## Diff. Exp {.tabset .tabset-fade .tabset-pills}

```{r warning=F,}
diffanal <- differential_abundance(MAE,
                            tax_level="genus",
                            input_da_condition=c("Class"),
                            input_da_condition_covariate = c("Sex", "imputed_smoking_label"),
                            min_num_filter  = 500,
                            input_da_padj_cutoff = 0.05, method = 'DESeq2')
DT::datatable(diffanal)
#write.xlsx(diffanal, file=file.path(PATH, "06_30_diffanal_microbes.xlsx"))

#get microbes unique to each contrast-condition
#up in first group and down in the other - up reg
microbe_list <- list("ctrl.vs.hknr_up"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 2-HkNR' & diffanal$padj <=0.05 & diffanal$log2FoldChange > 1)],
     "ctrl.vs.hknr_down"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 2-HkNR' & diffanal$padj <=0.05 & diffanal$log2FoldChange < -1)],
      "ctrl.vs.dys_up"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 3-Dysplasia' & diffanal$padj <=0.05 & diffanal$log2FoldChange > 1)],
     "ctrl.vs.dys_down"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 3-Dysplasia' & diffanal$padj <=0.05 & diffanal$log2FoldChange < -1)],
     "ctrl.vs.cancer_up"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 4-Cancer' & diffanal$padj <=0.05 & diffanal$log2FoldChange > 1)],
     "ctrl.vs.cancer_down"=diffanal$microbe[which(diffanal$Contrast=='1-Control vs. 4-Cancer' & diffanal$padj <=0.05 & diffanal$log2FoldChange < -1)])
microbe_list

#saveRDS(microbe_list, file.path("results/06_30_microbe_diffex_list.RDS"))
```

## Violin Plots {.tabset .tabset-fade .tabset-pills}

```{r}
volcano_wrapper <- function(df){
  ggplot(data=df, aes(x=log2FoldChange, y=-log10(padj), col=diffex, label=microbe)) +
        geom_point() + 
        theme_minimal() +
        geom_text_repel() +
        scale_color_manual(values=c("red", "blue")) +
        geom_vline(xintercept=c(-0.6, 0.6), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")
}
```

### Cancer vs Ctrl

```{r}
library(ggrepel)
df <- diffanal[which(diffanal$Contrast=='1-Control vs. 4-Cancer'),]
df$log2FoldChange <- -df$log2FoldChange
df$diffex <- ifelse(df$log2FoldChange >= 1 & df$padj <= 0.05, "Up", ifelse(df$log2FoldChange <= -1 & df$padj <= 0.05, "Down", "No diff"))
df$diffex <- factor(df$diffex, levels = c("Up", "Down")) 

g1 <- volcano_wrapper(df = df)
g1
```

### Dys vs Ctrl

```{r}
library(ggrepel)
df <- diffanal[which(diffanal$Contrast=='1-Control vs. 3-Dysplasia'),]
df$log2FoldChange <- -df$log2FoldChange
df$diffex <- ifelse(df$log2FoldChange >= 1 & df$padj <= 0.05, "Up", ifelse(df$log2FoldChange <= -1 & df$padj <= 0.05, "Down", "No diff"))
df$diffex <- factor(df$diffex, levels = c("Up", "Down")) 

g2 <- volcano_wrapper(df = df)
g2
```

### HkNR vs Ctrl

```{r}
library(ggrepel)
df <- diffanal[which(diffanal$Contrast=='1-Control vs. 2-HkNR'),]
df$log2FoldChange <- -df$log2FoldChange
df$diffex <- ifelse(df$log2FoldChange >= 1 & df$padj <= 0.05, "Up", ifelse(df$log2FoldChange <= -1 & df$padj <= 0.05, "Down", "No diff"))
df$diffex <- factor(df$diffex, levels = c("Up", "Down")) 

g3 <- volcano_wrapper(df = df)
g3
```



## MSEA {.tabset .tabset-fade .tabset-pills}

Microbe set enrichment analysis [(MSEA)](https://www.nature.com/articles/s41598-020-78511-y#Sec18)
Load results from msea - ran on scc with a jupyter notebook

### OSCC vs. Ctrl

```{r}
msea_cancer_up <- read.csv(file.path(PATH, "results/msea_cancer_ctrl_up.csv"))
msea_cancer_dn <- read.csv(file.path(PATH, "results/msea_cancer_ctrl_dn.csv"))
msea_pml_up <- read.csv(file.path(PATH, "results/msea_pml_ctrl_up.csv"))
msea_pml_dn <- read.csv(file.path(PATH, "results/msea_pml_ctrl_dn.csv"))

msea_cancer_up <- msea_cancer_up[msea_cancer_up$qvalue<=0.05,]
msea_cancer_dn <- msea_cancer_dn[msea_cancer_dn$qvalue<=0.05,]#none significant
msea_pml_up <- msea_pml_up[msea_pml_up$qvalue<=0.05,]
msea_pml_dn <- msea_pml_dn[msea_pml_dn$qvalue<=0.05,]#none significant

msea_genes <- list("cancer" = msea_cancer_up$term, 
                   "pml"=msea_pml_up$term)

HALLMARK <-  msigdb_gsets("Homo sapiens", "H", "")
names(HALLMARK$genesets) <- names(HALLMARK$genesets) %>% strsplit( "HALLMARK_" ) %>% sapply( tail, 1 )

REACTOME <- msigdb_gsets(species="Homo sapiens", category="C2", subcategory="CP:REACTOME", clean = TRUE)
names(REACTOME$genesets) <- names(REACTOME$genesets) %>% strsplit( "REACTOME_" ) %>% sapply( tail, 1 )


hyp_mic1 <- hypeR(signature = msea_genes, genesets = HALLMARK, background = 1300)
hyp_mic2 <- hypeR(signature = msea_genes, genesets = HALLMARK)
hyp_dots(hyp_mic1, merge = T, fdr = 0.1, title = "Hallmark(bck=1300)")
hyp_mic1$as.list()
hyp_dots(hyp_mic2, merge = T, fdr = 0.1, title = "Hallmark(bck=all)")
hyp_mic2$as.list()

hyp_mic3 <- hypeR(signature = msea_genes, genesets = REACTOME, background = 1300)
hyp_mic4 <- hypeR(signature = msea_genes, genesets = REACTOME)
hyp_dots(hyp_mic3, merge = T, fdr = 0.1, title = "Reactome(bck=1300)")
hyp_mic3$as.list()
hyp_dots(hyp_mic4, merge = T, fdr = 0.1, title = "Reactome(bck=all)")
hyp_mic4$as.list()

```


### Cancer genes

```{r}
diffanal_res <- xlsx::read.xlsx(file.path(PATH, "results/pml_diffex_wo_infl/PML.DiffEx.results.Can.vs.Ctrl.xlsx"), sheetIndex = 1)
diffanal_res <- diffanal_res[diffanal_res$padj<=0.05,]
diff_cancer <- diffanal_res[diffanal_res$gene %in% msea_genes$cancer, ]

ggplot(data = diff_cancer, aes(x=gene, y = log2FoldChange),)+
  geom_bar(aes( fill=log2FoldChange>0), stat = "identity")+ 
  theme_bw()+  
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(guide = "none", breaks = c(TRUE, FALSE), values=c("blue4", "red"))
```


### Cancer vs Ctrl up

```{r}
library(tidyverse)
msea_cancer_up_new<- separate_rows(msea_cancer_up, shared, sep = ",", convert = T)
msea_cancer_up_new$shared <- gsub(msea_cancer_up_new$shared, pattern = "\\[", replacement = "")
msea_cancer_up_new$shared <- gsub(msea_cancer_up_new$shared, pattern = "\\]", replacement = "")
msea_cancer_up_new$shared <- gsub(msea_cancer_up_new$shared, pattern = "\\'", replacement = "")
msea_cancer_up_new$shared <- gsub(msea_cancer_up_new$shared, pattern = " ", replacement = "")
msea_cancer_up_new

```

### igraph

```{r fig.height=10}

tdf <- table(msea_cancer_up_new[, c('term', 'shared')])
g <- graph.incidence(tdf, weighted = TRUE)
is.bipartite(g)
deg <- degree(g, mode="all")
deg <- ifelse(names(deg) %in% msea_cancer_up_new$term, deg*5, deg)

colrs <- c("green", "cyan")[V(g)$type + 1L]
lay <- layout_as_bipartite(g)[,2:1]

V(g)$shape <- ifelse(V(g) %in% msea_cancer_up_new$term, "square", "circle")

plot(g, 
     vertex.color = colrs, 
     layout = lay, 
     vertex.size=deg,
     asp = 5,
     vertex.size=c(15:24))

```

### Final igraph
- Choose genes that come up diff analysis
- to adjust for degree size for genes double it

```{r fig.height=10}
msea_cancer_up_de <- msea_cancer_up_new[msea_cancer_up_new$term %in% diff_cancer$gene, ]
df <- msea_cancer_up_de[, c("term", "shared", "combined_score")]
transformed_mat <- xtabs(combined_score~., df)
attr(transformed_mat, "class") <- NULL #ignore this

g <- graph.incidence(transformed_mat, weighted = TRUE)
is.bipartite(g)
deg <- degree(g, mode="all")
colrs <- c("#FFD580","#6CA6CD")[V(g)$type + 1L]


V(g)$shape <- ifelse(V(g)$name %in% unquote(msea_cancer_up_de$term), "circle", "square")

LO = matrix(0, nrow=vcount(g), ncol=2)
LO[!V(g)$type, 2] = 1

LO[V(g)$type, 1]  = rank(V(g)$name[V(g)$type]) - 1
LO[!V(g)$type, 1] = (rank(V(g)$name[!V(g)$type]) - 1) * 
    (sum(V(g)$type) - 1)  /  (sum(!V(g)$type) - 1)

#for a vertical bipartite graph
LO <- LO[,2:1]


plot(g, 
     vertex.color = adjustcolor(colrs,  alpha.f = .6), 
     vertex.frame.color = colrs,
     layout = LO, 
     vertex.size = ifelse(V(g)$name %in% unquote(msea_cancer_up_de$term), deg*2, deg*1.95),
     #vertex.size = deg*3,
     vertex.label.dist = 0,
     vertex.label.cex=0.3,
     vertex.frame.width=0.1,
     edge.width=abs(E(g)$weight)/200, 
     vertex.label.family="Arial",
     edge.color=ifelse(E(g)$weight > 0, "red","blue"),
     asp = 2.5)
```


```{r}
microbes_cancer_shared <- unique(msea_cancer_up_de$shared)
DT::datatable(msea_cancer_up_de)
```


### PML vs. Ctrl

```{r}
diffanal_res_hknr <- xlsx::read.xlsx(file.path(PATH, "results/pml_diffex_wo_infl/PML.DiffEx.results.Hknr.vs.Ctrl.xlsx"), sheetIndex = 1)
diffanal_res_dys <- xlsx::read.xlsx(file.path(PATH, "results/pml_diffex_wo_infl/PML.DiffEx.results.Dys.vs.Ctrl.xlsx"), sheetIndex = 1)

diffanal_res_hknr$gene <- diffanal_res_hknr$NA.
diffanal_res_dys$gene <- diffanal_res_dys$NA.

diffanal_pml <- rbind(diffanal_res_dys, diffanal_res_hknr)

diffanal_pml <- diffanal_pml[diffanal_pml$padj<=0.05,]
diff_pml<- diffanal_pml[diffanal_pml$gene %in% msea_genes$pml, ]

ggplot(data = diff_pml, aes(x=gene, y = log2FoldChange),)+
  geom_bar(aes( fill=log2FoldChange>0), stat = "identity")+ 
  theme_bw()+  
  ggplot2::theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(guide = "none", breaks = c(TRUE, FALSE), values=c("blue4", "red"))
```

### PML vs Ctrl up

```{r}
msea_pml_up_new<- separate_rows(msea_pml_up, shared, sep = ",", convert = T)
msea_pml_up_new$shared <- gsub(msea_pml_up_new$shared, pattern = "\\[", replacement = "")
msea_pml_up_new$shared <- gsub(msea_pml_up_new$shared, pattern = "\\]", replacement = "")
msea_pml_up_new$shared <- gsub(msea_pml_up_new$shared, pattern = "\\'", replacement = "")
msea_pml_up_new$shared <- gsub(msea_pml_up_new$shared, pattern = " ", replacement = "")
msea_pml_up_new

```

- Choose genes that come up diff analysis
- to adjust for degree size for genes double it


```{r}
msea_pml_up_de <- msea_pml_up_new[msea_pml_up_new$term %in% diff_pml$gene, ]
df <- msea_pml_up_de[, c("shared", "term", "combined_score")]
transformed_mat <- xtabs(combined_score~., df)
attr(transformed_mat, "class") <- NULL #ignore this

g <- graph.incidence(transformed_mat, weighted = TRUE)
is.bipartite(g)
deg <- degree(g, mode="all")
colrs <- c("#7EC384", "#FFD580")[V(g)$type + 1L]

V(g)$shape <- ifelse(V(g)$name %in% unquote(msea_pml_up_de$term), "circle", "square")
#V(g)$label.cez <- 1

LO = matrix(0, nrow=vcount(g), ncol=2)
LO[!V(g)$type, 2] = 1

LO[V(g)$type, 1]  = rank(V(g)$name[V(g)$type]) - 1
LO[!V(g)$type, 1] = (rank(V(g)$name[!V(g)$type]) - 1) * (sum(V(g)$type) - 1)  /  (sum(!V(g)$type) - 1)

#for a vertical bipartite graph
LO <- LO[,2:1]
#LO[,2] <- LO[,2]*2

plot(g, 
     vertex.color = adjustcolor(colrs,  alpha.f = .6), 
     vertex.frame.color = colrs,
     layout = LO, 
     vertex.size = ifelse(V(g)$name %in% unquote(msea_pml_up_de$term), deg*1.25, deg*1.25),
     vertex.label.dist = ifelse(V(g)$name %in% unquote(msea_pml_up_de$term), 0, 0),
     vertex.label.cex= ifelse(V(g)$name %in% unquote(msea_pml_up_de$term), 0.18, 0.35),
     vertex.frame.width=0.1,
     vertex.label.font=3,
     edge.width=abs(E(g)$weight)/200, 
     vertex.label.family="Arial",
     edge.color=ifelse(E(g)$weight > 0, "red", "blue"),  asp = 3)

```




